---
title: "Synchrony_Dissertation_Process"
author: "Ester Kirby"
date: "2026-06-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synchrony Dissertation Process

This is an RMarkdown serving as a data management plan and lab book for a dissertation undertaken in partial fulfillment of an MSc in Ecology, Evolution and Biodiversity (2026). 

``` {r, echo=FALSE}
library(lme4)
library(insight)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(viridis)
library(rstan)
library(patchwork)
```

```{r Importing and Cleaning Data, echo=FALSE}
#importing bird phenology dataframe
phendat <- read.csv("data/Bird_Phenology.csv")

#importing site details dataframe
sitedat <- read.csv("data/Site_Details.csv")

#importing adults dataframe
adults <- read.csv("data/Adults.csv")
adults <- adults[adults$season == 'spring',]

#importing nestlings dataframe
nestlings <- read.csv("data/Nestlings.csv")

#changing -999s (0s in fledgling success caused by predation) to 0s
phendat$suc[phendat$suc == "-999"] <- 0

#checking for duplicates
phendat$ID <- as.factor(paste(phendat$year, phendat$site, phendat$box, sep = "_")) 
length(phendat$ID)
length(unique(phendat$ID))
phendat$ID[duplicated(phendat$ID)]
#2230 entries, 5 of which not unique

#keeping just the last brood (completed)
phendat <- phendat[!duplicated(phendat$ID, fromLast = TRUE), ]

phendat$ID[duplicated(phendat$ID)]
#success!

#removing non focal species
bludat <- phendat[phendat$species == "bluti",]
```

Above, I imported the relevant dataframes from the phenoweb project. 
I then changed -999s in column 'suc' to 0s in phendat as these indicated 0 success due to predation. This is because predation is impacted by and has impacts on synchrony, and therefore is biologically significant and does not need to be flagged as anomalous. The only concern is if the nestboxes cause a change in predation probability from natural levels - a literature search will be conducted to advise on this. 
I then identified duplicates, of which there were only 5. These were caused by birds abandoning nests and then new broods taking. I removed the first broods from the data as they did not represent proper brood attempts.
Finally, I created a separate dataframe containing only rows representing occupation by blue tits. 

```{r clutch swap treatment, echo=FALSE}
bludatswap <- bludat %>%
  mutate(clutch.swap.treatment = ifelse(clutch.swap.treatment %in% 1:9, "swap", clutch.swap.treatment))

bludatswap <- bludatswap[bludatswap$year %in% c("2017", "2018", "2019"), ]

bludatswap$clutch.swap.treatment <- substr(bludatswap$clutch.swap.treatment, 1, 3)

clutchswapmod <- lm(suc ~ clutch.swap.treatment, data = bludatswap)
summary(clutchswapmod)

swapmodmixed <- lmer(suc ~ clutch.swap.treatment + (1|site) + (1|year), data = bludatswap)
summary(swapmodmixed)

#relevel to make unm the reference
str(bludatswap)
bludatswap$clutch.swap.treatment <- as.factor(bludatswap$clutch.swap.treatment)
bludatswap$clutch.swap.treatment <- relevel(bludatswap$clutch.swap.treatment, ref = "unm")

#making year a factor to reduce noise
bludatswap$year <- as.factor(bludatswap$year)

#removing nas in swap
bludatswap <- bludatswap %>%
  mutate(clutch.swap.treatment = na_if(clutch.swap.treatment, "")) %>%
  drop_na(clutch.swap.treatment)

#popping odd and not possible together 
bludatswap <- bludat %>%
  mutate(clutch.swap.treatment = as.character(clutch.swap.treatment)) %>%
  mutate(clutch.swap.treatment = ifelse(clutch.swap.treatment == "not", "odd", clutch.swap.treatment))

swapmodmixed <- lmer(suc ~ (1|clutch.swap.treatment) + year + (1|site), data = bludatswap)
summary(swapmodmixed)

tapply(bludatswap$clutch.swap.treatment, bludatswap$clutch.swap.treatment, length)
tapply(bludatswap$suc, bludatswap$clutch.swap.treatment, mean, na.rm=TRUE)

```

Above, I isolated the clutch swap experiment years and ran a linear mixed model to look for variance attributed to clutch swap treatment as opposed to unmanipulated clutches. This allowed me to decide whether to include these years, just the unmanipulated broods, or exclude entirely. Overall variance attributed to treatment was low compared to residual, and mean success was similar between unmanipulated and swapped clutches. Therefore, it appears the effect of the clutch swap experiment on fledgling success was negligible and these years can be included in full. 

```{r fledgling success initial glmm, echo=FALSE}

#very early model attempt on fledgling success
bludat$site_year <- as.factor(paste(bludat$site, bludat$year, sep = "_")) 
sucmod <- lmer(suc ~ (1|year) + (1|site) + (1|site_year), data = bludat) 
summary(sucmod)

get_variance_intercept(sucmod)

#extracting the variance components 
var_components <- VarCorr(sucmod)
var_intercept_site_year <- attr(var_components$site_year, "stddev")^2
var_intercept_site <- attr(var_components$site, "stddev")^2
var_intercept_year <- attr(var_components$year, "stddev")^2

#a crude look at ICC synchrony from the model 
sucsync <- var_intercept_year/(var_intercept_year + var_intercept_site_year)

sucsync
```

Here, I ran a simple linear mixed model on fledgling success with year, site and site year random effects. I then extracted the variance components and calculated the ICC of fledgling success synchrony

##Hoopoe Integrated Population Model from Kery and Schaub 2012
#Translated to Rstan by Sanjana 

```{r Sanjanas code, echo=FALSE}
# Simulate datasets for two more sites
nsites <- 3
nyears <- 9

# Note: CJS model is specified as a multinomial likelihood (i.e. individual capture histories are collapsed into m-arrays) and were not specified as individual-level state-space formulation I had done previously for blue tits
# More details on m-arrays and code to convert individual capture histories to m-arrays can be found in Chapter 7.10 in book, "Bayesian Population Analysis using WinBUGS --- A Hierarchical Perspective" (2012) by Marc Kéry and Michael Schaub.

# All site-1 associated data is the hoopoe dataset from Chapter 11, Kery & Schaub 2012 book
# Capture-recapture data: m-array of juveniles and adults (these are males and females together)
marray_j_site1 <- structure(
  c(15, 0, 0, 0, 0, 0, 0, 0, 3, 34, 0, 0, 0, 0, 0, 0,
    0, 9, 56, 0, 0, 0, 0, 0, 0, 1, 8, 48, 0, 0, 0, 0, 0, 0, 1, 3,
    45, 0, 0, 0, 0, 0, 0, 1, 13, 27, 0, 0, 0, 0, 0, 0, 2, 7, 37,
    0, 0, 0, 0, 0, 0, 0, 3, 39, 198, 287, 455, 518, 463, 493, 434, 405),
  .Dim = 8:9)

marray_a_site1 <- structure(
  c(14, 0, 0, 0, 0, 0, 0, 0, 2, 22, 0, 0, 0, 0, 0, 0,
    0, 4, 34, 0, 0, 0, 0, 0, 0, 0, 2, 51, 0, 0, 0, 0, 0, 0, 0, 3,
    45, 0, 0, 0, 0, 0, 0, 0, 3, 44, 0, 0, 0, 0, 0, 0, 0, 3, 48, 0,
    0, 0, 0, 0, 0, 0, 2, 51, 43, 44, 79, 94, 118, 113, 99, 90),
  .Dim = 8:9)

y_site1 <- c(32, 42, 64, 85, 82, 78, 73, 69, 79) # Population count data  (the maximal number of simultaneously occupied nest boxes in each year)
J_site1 <- c(189, 274, 398, 538, 520, 476, 463, 438) # Number of offspring
R_site1 <- c(28, 36, 57, 77, 81, 83, 77, 72) # Number of surveyed broods

# Site 2 
set.seed(42)
y_site2 <- round(y_site1 * 0.60 * exp(rnorm(nyears, 0, 0.10)))
J_site2 <- round(J_site1 * 0.60 * exp(rnorm(nyears - 1, 0, 0.10)))
R_site2 <- round(R_site1 * 0.60 * exp(rnorm(nyears - 1, 0, 0.10)))

marray_j_site2 <- structure(
  c(12, 0, 0, 0, 0, 0, 0, 0, 2, 28, 0, 0, 0, 0, 0, 0,
    0, 7, 45, 0, 0, 0, 0, 0, 0, 2, 6, 38, 0, 0, 0, 0, 0, 0, 2, 4,
    36, 0, 0, 0, 0, 0, 0, 1, 10, 22, 0, 0, 0, 0, 0, 0, 1, 5, 29,
    0, 0, 0, 0, 0, 0, 0, 2, 31, 210, 265, 430, 490, 440, 470, 410, 385),
  .Dim = 8:9)

marray_a_site2 <- structure(
  c(11, 0, 0, 0, 0, 0, 0, 0, 3, 19, 0, 0, 0, 0, 0, 0,
    0, 3, 29, 0, 0, 0, 0, 0, 0, 0, 3, 44, 0, 0, 0, 0, 0, 0, 0, 2,
    38, 0, 0, 0, 0, 0, 0, 0, 2, 38, 0, 0, 0, 0, 0, 0, 0, 2, 41, 0,
    0, 0, 0, 0, 0, 0, 1, 44, 40, 38, 68, 82, 105, 100, 88, 78),
  .Dim = 8:9)

# Site 3 
set.seed(43)
y_site3 <- round(y_site1 * 1.20 * exp(rnorm(nyears, 0, 0.12)))
J_site3 <- round(J_site1 * 1.20 * exp(rnorm(nyears - 1, 0, 0.12)))
R_site3 <- round(R_site1 * 1.20 * exp(rnorm(nyears - 1, 0, 0.12)))

marray_j_site3 <- structure(
  c(18, 0, 0, 0, 0, 0, 0, 0, 4, 40, 0, 0, 0, 0, 0, 0,
    0, 11, 63, 0, 0, 0, 0, 0, 0, 1, 10, 55, 0, 0, 0, 0, 0, 0, 1, 4,
    52, 0, 0, 0, 0, 0, 0, 2, 15, 31, 0, 0, 0, 0, 0, 0, 2, 9, 43,
    0, 0, 0, 0, 0, 0, 0, 4, 45, 185, 305, 478, 540, 480, 510, 452, 420),
  .Dim = 8:9)

marray_a_site3 <- structure(
  c(16, 0, 0, 0, 0, 0, 0, 0, 2, 25, 0, 0, 0, 0, 0, 0,
    0, 5, 39, 0, 0, 0, 0, 0, 0, 0, 2, 58, 0, 0, 0, 0, 0, 0, 0, 4,
    51, 0, 0, 0, 0, 0, 0, 0, 4, 50, 0, 0, 0, 0, 0, 0, 0, 4, 54, 0,
    0, 0, 0, 0, 0, 0, 3, 57, 46, 50, 88, 102, 128, 120, 108, 97),
  .Dim = 8:9)

# Combine data from all sites
# Combining pop counts
(y_mat <- rbind(y_site1, y_site2, y_site3))

# Combining offpsring numbers (J) and surveyed broods (R)
# J, R: [nsites, nyears-1] format
(J_mat <- rbind(J_site1, J_site2, J_site3))
(R_mat <- rbind(R_site1, R_site2, R_site3))

# Combining recapture histories (m-array format)
# marray: [nsites, nyears-1, nyears] format
marray_j_3d <- array(NA_integer_, dim = c(nsites, nyears - 1, nyears))
marray_a_3d <- array(NA_integer_, dim = c(nsites, nyears - 1, nyears))

marray_j_3d[1,,] <- marray_j_site1
marray_j_3d[2,,] <- marray_j_site2
marray_j_3d[3,,] <- marray_j_site3

marray_a_3d[1,,] <- marray_a_site1
marray_a_3d[2,,] <- marray_a_site2
marray_a_3d[3,,] <- marray_a_site3

# Put together all data to feed into stan model
stan_data <- list(
  nyears    = nyears,
  nsites    = nsites,
  y         = y_mat,      
  J         = J_mat,      
  R         = R_mat,        
  marray_j  = marray_j_3d, 
  marray_a  = marray_a_3d  
)

# Give Stan parameters to estimate
params <- c(
  "phij", "phia", "f", "omega", "p",
  "mphij", "mphia", "mfec", "mim",
  "lambda", "mlam",
  "sig_phij",      "sig_phia",      "sig_fec",      "sig_im",
  "sig_site_phij", "sig_site_phia", "sig_site_fec", "sig_site_im",
  "sig_sy_phij",   "sig_sy_phia",   "sig_sy_fec",   "sig_sy_im",
  "N1", "NadSurv", "Nadimm", "Ntot",
  "var_phij_year", "var_phij_siteyear", "icc_phij",
  "var_phia_year", "var_phia_siteyear", "icc_phia",
  "var_fec_year", "var_fec_siteyear", "icc_fec",
  "var_im_year", "var_im_siteyear", "icc_im"
)

# Setting stan model options
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
set.seed(123)

# MCMC settings
ni <- 20000   
nb <- 10000   
nt <- 5       
nc <- 4

## Initial values
inits <- lapply(1:nc, function(i) {
  list(l_mphij = rnorm(1, 0.2, 0.5),
       l_mphia = rnorm(1, 0.2, 0.5),
       l_mfec = rnorm(1, 0.2, 0.5),
       l_mim = rnorm(1, 0.2, 0.5),
       l_p = rnorm(1, 0.2, 1),
       sig_phij = runif(1, 0.1, 10),
       sig_phia = runif(1, 0.1, 10),
       sig_fec = runif(1, 0.1, 10),
       sig_im = runif(1, 0.1, 10),
       sig_site_phij = runif(1, 0.1, 10),
       sig_site_phia = runif(1, 0.1, 10),
       sig_site_fec = runif(1, 0.1, 10),
       sig_site_im = runif(1, 0.1, 10),
       sig_sy_phij = runif(1, 0.1, 10),
       sig_sy_phia = runif(1, 0.1, 10),
       sig_sy_fec = runif(1, 0.1, 10),
       sig_sy_im = runif(1, 0.1, 10),
       N1      = lapply(1:nsites, function(s) round(runif(nyears,  1, 50))),
       NadSurv = lapply(1:nsites, function(s) round(runif(nyears,  5, 50))),
       Nadimm  = lapply(1:nsites, function(s) round(runif(nyears,  1, 50)))
)})

# Run stan model
#ipm_hoopoe <- stan(
#  file    = "ipm_hoopoe.stan",
#  data    = stan_data,
#  init    = inits,
#  pars    = params,
#  chains  = nc,
#  iter    = ni,
#  warmup  = nb,
#  thin    = nt,
#  seed    = 2,
#  control = list(
#    adapt_delta   = 0.99,
#    max_treedepth = 15    
#  )
#)

rds_file_path <- "ipm_hoopoe.rds"

saveRDS(ipm_hoopoe, rds_file_path)

ipm_hoopoe <- readRDS("ipm_hoopoe.rds")

# Print stan model output
print(ipm_hoopoe,
      pars = c("mphij", "mphia", "mfec", "mim",
               "sig_phij",      "sig_phia",      "sig_fec",      "sig_im",
               "sig_site_phij", "sig_site_phia", "sig_site_fec", "sig_site_im",
               "sig_sy_phij",   "sig_sy_phia",   "sig_sy_fec",   "sig_sy_im",
               "mlam",   "var_phij_year", "var_phij_siteyear", "icc_phij",
               "var_phia_year", "var_phia_siteyear", "icc_phia",
               "var_fec_year", "var_fec_siteyear", "icc_fec",
               "var_im_year", "var_im_siteyear", "icc_im"),
      digits = 3)

# Get posterior distributions
posterior_ipm <- as.data.frame(ipm_hoopoe)

# extract summary (mean + 95% CrI) for parameters
extract_summary <- function(posterior, pattern) {
  cols <- grep(pattern, names(posterior), value = TRUE)
  data.frame(
    param = cols,
    mean  = colMeans(posterior[, cols, drop = FALSE]),
    lower = apply(posterior[, cols, drop = FALSE], 2, quantile, 0.025),
    upper = apply(posterior[, cols, drop = FALSE], 2, quantile, 0.975)
  )
}

# Get nyears and nsites
nyears  <- stan_data$nyears
nsites  <- stan_data$nsites
years   <- 1:nyears
years_m1 <- 1:(nyears - 1)  # for parameters indexed over nyears-1

site_labels <- paste0("Site ", 1:nsites)

# Estimated Ntot per site per year
ntot_list <- lapply(1:nsites, function(s) {
  pattern <- paste0("Ntot\\[", s, ",")
  df <- extract_summary(posterior_ipm, pattern)
  df$year <- years
  df$site <- site_labels[s]
  df
})
ntot_df <- bind_rows(ntot_list)

# Observed counts per site per year
obs_counts <- expand.grid(site = site_labels, year = years) %>%
  arrange(site, year) %>%
  mutate(observed = as.vector(t(stan_data$y)))  # y is [nsites, nyears]

# Merge both dfs
ntot_df <- left_join(ntot_df, obs_counts, by = c("site", "year"))

# Plot
p_ntot <- ggplot(ntot_df, aes(x = year)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = mean, colour = "Estimated Ntot"), linewidth = 0.9) +
  geom_point(aes(y = mean, colour = "Estimated Ntot"), size = 1.5) +
  geom_line(aes(y = observed, colour = "Observed count"),
            linewidth = 0.9, linetype = "dashed") +
  geom_point(aes(y = observed, colour = "Observed count"), size = 1.5) +
  scale_colour_manual(values = c("Estimated Ntot" = "steelblue",
                                 "Observed count"  = "firebrick")) +
  facet_wrap(~ site, scales = "free_y") +
  labs(title = "Population size", x = "Year", y = "N",
       colour = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom") +
  scale_x_continuous(n.breaks=8)

p_ntot

# Exercises for Ester -

# Exercise1: Plot and investigate temporal trends in other demographic parameters (survival, productivity, immigration, population growth rate etc) across the different sites (doing this will be handy for blue tit data analysis)

#productivity/fecundity 
# Estimated f per site per year
f_list <- lapply(1:nsites, function(s) {
  pattern <- paste0("f\\[", s, ",")
  df <- extract_summary(posterior_ipm, pattern)
  df$year <- years_m1
  df$site <- site_labels[s]
  df
})
f_est <- bind_rows(f_list)

# Observed f per site per year
obs_f <- expand.grid(site = site_labels, year = years_m1) %>%
  arrange(site, year) %>%
  mutate(observed = as.vector(t(stan_data$J / stan_data$R)))

# Merge both dfs
f_df <- left_join(f_est, obs_f, by = c("site", "year"))

# Plot
p_f <- ggplot(f_df, aes(x = year)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = mean, colour = "Estimated f"), linewidth = 0.9) +
  geom_point(aes(y = mean, colour = "Estimated f"), size = 1.5) +
  geom_line(aes(y = observed, colour = "Observed f"),
            linewidth = 0.9, linetype = "dashed") +
  geom_point(aes(y = observed, colour = "Observed f"), size = 1.5) +
  scale_colour_manual(values = c("Estimated f" = "steelblue",
                                 "Observed f"  = "firebrick")) +
  facet_wrap(~ site, scales = "free_y") +
  labs(title = "Fecundity per Nest", x = "Year", y = "N Fledglings/female",
       colour = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom") +
  scale_x_continuous(n.breaks=8)

#p_f
```

```{r hoopoe plotting - Sanjanas exercises, echo=FALSE}
# Exercise1: Plot and investigate temporal trends in other demographic parameters (survival, productivity, immigration, population growth rate etc) across the different sites (doing this will be handy for blue tit data analysis)

#productivity/fecundity 
# Estimated f per site per year
f_list <- lapply(1:nsites, function(s) {
  pattern <- paste0("f\\[", s, ",")
  df <- extract_summary(posterior_ipm, pattern)
  df$year <- years_m1
  df$site <- site_labels[s]
  df
})
f_est <- bind_rows(f_list)

# Observed f per site per year
obs_f <- expand.grid(site = site_labels, year = years_m1) %>%
  arrange(site, year) %>%
  mutate(observed = as.vector(t(stan_data$J / stan_data$R)))

# Merge both dfs
f_df <- left_join(f_est, obs_f, by = c("site", "year"))

# Plot
p_f <- ggplot(f_df, aes(x = year)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = mean, colour = "Estimated f"), linewidth = 0.9) +
  geom_point(aes(y = mean, colour = "Estimated f"), size = 1.5) +
  geom_line(aes(y = observed, colour = "Observed f"),
            linewidth = 0.9, linetype = "dashed") +
  geom_point(aes(y = observed, colour = "Observed f"), size = 1.5) +
  scale_colour_manual(values = c("Estimated f" = "steelblue",
                                 "Observed f"  = "firebrick")) +
  facet_wrap(~ site, scales = "free_y") +
  labs(title = "Fecundity per Nest", x = "Year", y = "N Fledglings/female",
       colour = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom") +
  scale_x_continuous(n.breaks=8)

#p_f


#survival
#juveniles
# Estimated phij per site per year
phij_list <- lapply(1:nsites, function(s) {
  pattern <- paste0("phij\\[", s, ",")
  df <- extract_summary(posterior_ipm, pattern)
  df$year <- years_m1
  df$site <- site_labels[s]
  df
})
phij_est <- bind_rows(phij_list)

#adults
phia_list <- lapply(1:nsites, function(s) {
  pattern <- paste0("phia\\[", s, ",")
  df <- extract_summary(posterior_ipm, pattern)
  df$year <- years_m1
  df$site <- site_labels[s]
  df
})
phia_est <- bind_rows(phia_list)


# Observed f per site per year
#obs_phij <- expand.grid(site = site_labels, year = years_m1) %>%
 # arrange(site, year) %>%
  #mutate(observed = as.vector(t()))

# Merge both dfs
phi_df <- left_join(phij_est, phia_est, by = c("site", "year"))

# Plot
p_phi <- ggplot(phi_df, aes(x = year)) +
  geom_ribbon(aes(ymin = lower.x, ymax = upper.x), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = mean.x, colour = "Estimated juvenile survival"), linewidth = 0.9) +
  geom_point(aes(y = mean.x, colour = "Estimated juvenile survival"), size = 1.5) +
  geom_ribbon(aes(ymin = lower.y, ymax = upper.y), fill = "firebrick", alpha = 0.25) +
  geom_line(aes(y = mean.y, colour = "Estimated adult survival"), linewidth = 0.9) +
  geom_point(aes(y = mean.y, colour = "Estimated adult survival"), size = 1.5) +
  scale_colour_manual(values = c("Estimated juvenile survival" = "steelblue", "Estimated adult survival" = "firebrick")) +
  facet_wrap(~ site, scales = "free_y") +
  labs(title = "Estimated Survival", x = "Year", y = "Phi",
       colour = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom") +
  scale_x_continuous(n.breaks=8)

#p_phi

#population growth rate

lambda_list <- lapply(1:nsites, function(s) {
  pattern <- paste0("lambda\\[", s, ",")
  df <- extract_summary(posterior_ipm, pattern)
  df$year <- years_m1
  df$site <- site_labels[s]
  df
})
lambda_df <- bind_rows(lambda_list)

# Observed counts per site per year
#obs_counts <- expand.grid(site = site_labels, year = years) %>%
 # arrange(site, year) %>%
  #mutate(observed = as.vector(t(stan_data$y)))  # y is [nsites, nyears]

# Merge both dfs
#lambda_df <- left_join(lambda_df, obs_counts, by = c("site", "year"))

# Plot
p_lambda <- ggplot(lambda_df, aes(x = year)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = mean, colour = "Estimated lambda"), linewidth = 0.9) +
  geom_point(aes(y = mean, colour = "Estimated lambda"), size = 1.5) +
  #geom_line(aes(y = observed, colour = "Observed count"),
   #         linewidth = 0.9, linetype = "dashed") +
  #geom_point(aes(y = observed, colour = "Observed count"), size = 1.5) +
  scale_colour_manual(values = c("Estimated lambda" = "steelblue")) +
  facet_wrap(~ site, scales = "free_y") +
  labs(title = "Population growth rate", x = "Year", y = "Lambda",
       colour = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom") +
  scale_x_continuous(n.breaks=8)

#p_lambda

#immigration
Nadimm_list <- lapply(1:nsites, function(s) {
  pattern <- paste0("Nadimm\\[", s, ",")
  df <- extract_summary(posterior_ipm, pattern)
  df$year <- years
  df$site <- site_labels[s]
  df
})
Nadimm_df <- bind_rows(Nadimm_list)

# Observed counts per site per year
obs_counts <- expand.grid(site = site_labels, year = years) %>%
  arrange(site, year) %>%
  mutate(observed = as.vector(t(stan_data$y)))  # y is [nsites, nyears]

# Merge both dfs
Nadimm_df <- left_join(Nadimm_df, obs_counts, by = c("site", "year"))

# Plot
p_Nadimm <- ggplot(Nadimm_df, aes(x = year)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = mean, colour = "Estimated Nadimm"), linewidth = 0.9) +
  geom_point(aes(y = mean, colour = "Estimated Nadimm"), size = 1.5) +
#  geom_line(aes(y = observed, colour = "Observed count"),
#            linewidth = 0.9, linetype = "dashed") +
#  geom_point(aes(y = observed, colour = "Observed count"), size = 1.5) +
  scale_colour_manual(values = c("Estimated Nadimm" = "steelblue")) +
  facet_wrap(~ site, scales = "free_y") +
  labs(title = "N Immigrants", x = "Year", y = "N",
       colour = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom") +
  scale_x_continuous(n.breaks=8)

#p_Nadimm

rates_plots <- (p_ntot | p_lambda | p_f) / (p_phi | p_Nadimm)

rates_plots

# Exercise2: Plot and investigate posterior distributions of random effect variances and spatial synchrony (ICCs) for all demographic parameters (doing this will be handy for blue tit data analysis)

#lambda
lambda_pds <- posterior_ipm %>%
  select(contains("lambda")) %>%
  pivot_longer(cols = everything(), values_to = "Value")

pds_lambda <- ggplot(lambda_pds, aes(x = Value)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Pooled Posterior Distribution of Lambda",
       x = "Lambda Value",
       y = "Density") +
  theme_minimal()

#phia icc

phia_icc <- posterior_ipm %>%
  select(contains("icc_phia")) %>%
  pivot_longer(cols = everything(), values_to = "Value")

icc_phia_p <- ggplot(phia_icc, aes(x = Value)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Adult Survival ICC",
       x = "phia ICC",
       y = "Density") +
  theme_minimal()

#icc_phia_p

#var_phia_siteyear

var_phia_siteyear <- posterior_ipm %>%
  select(contains("var_phia_siteyear")) %>%
  pivot_longer(cols = everything(), values_to = "Value")

var_phia_siteyear_p <- ggplot(var_phia_siteyear, aes(x = Value)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Adult Survival Random Effect Variances",
       x = "phia var",
       y = "Density") +
  theme_minimal()

#var_phia_siteyear_p

#phij icc

phij_icc <- posterior_ipm %>%
  select(contains("icc_phij")) %>%
  pivot_longer(cols = everything(), values_to = "Value")

icc_phij_p <- ggplot(phij_icc, aes(x = Value)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Juvenile Survival ICC",
       x = "phij ICC",
       y = "Density") +
  theme_minimal()

#icc_phij_p

#var_phij_siteyear

var_phij_siteyear <- posterior_ipm %>%
  select(contains("var_phij_siteyear")) %>%
  pivot_longer(cols = everything(), values_to = "Value")

var_phij_siteyear_p <- ggplot(var_phij_siteyear, aes(x = Value)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Juvenile Survival Random Effect Variances",
       x = "phij var",
       y = "Density") +
  theme_minimal()

#var_phij_siteyear_p

#fec icc

fec_icc <- posterior_ipm %>%
  select(contains("icc_fec")) %>%
  pivot_longer(cols = everything(), values_to = "Value")

icc_fec_p <- ggplot(fec_icc, aes(x = Value)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Fecundity ICC",
       x = "fec ICC",
       y = "Density") +
  theme_minimal()

#icc_fec_p

#var_fec_siteyear

var_fec_siteyear <- posterior_ipm %>%
  select(contains("var_fec_siteyear")) %>%
  pivot_longer(cols = everything(), values_to = "Value")

var_fec_siteyear_p <- ggplot(var_fec_siteyear, aes(x = Value)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Fecundity Random Effect Variances",
       x = "fec var",
       y = "Density") +
  theme_minimal()

#var_fec_siteyear_p

#im icc

im_icc <- posterior_ipm %>%
  select(contains("icc_im")) %>%
  pivot_longer(cols = everything(), values_to = "Value")

icc_im_p <- ggplot(im_icc, aes(x = Value)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Immigration ICC",
       x = "im ICC",
       y = "Density") +
  theme_minimal()

#icc_im_p

#var_im_siteyear

var_im_siteyear <- posterior_ipm %>%
  select(contains("var_im_siteyear")) %>%
  pivot_longer(cols = everything(), values_to = "Value")

var_im_siteyear_p <- ggplot(var_im_siteyear, aes(x = Value)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Immigration Random Effect Variances",
       x = "im var",
       y = "Density") +
  theme_minimal()

#var_im_siteyear_p

icc_var_plots <- (var_phia_siteyear_p | icc_phia_p) / (var_phij_siteyear_p | icc_phij_p) / (var_fec_siteyear_p | icc_fec_p) / (var_im_siteyear_p | icc_im_p)

icc_var_plots
```

``` {r marray code, echo=FALSE}
# Function to create a m-array based on capture-histories (CH)
marray <- function(CH){
  nind <- dim(CH)[1]
  n.occasions <- dim(CH)[2]
  m.array <- matrix(data = 0, ncol = n.occasions+1, nrow = n.occasions)
  # Calculate the number of released individuals at each time period
  for (t in 1:n.occasions){
    m.array[t,1] <- sum(CH[,t])
  }
  for (i in 1:nind){
    pos <- which(CH[i,]!=0)
    g <- length(pos)
    for (z in 1:(g-1)){
      m.array[pos[z],pos[z+1]] <- m.array[pos[z],pos[z+1]] + 1
    } #z
  } #i
  # Calculate the number of individuals that is never recaptured
  for (t in 1:n.occasions){
    m.array[t,n.occasions+1] <- m.array[t,1] - sum(m.array[t,2:n.occasions])
  }
  out <- m.array[1:(n.occasions-1),2:(n.occasions+1)]
  return(out)
}
```

``` {r ALN M-Array, echo=FALSE}

ALN_adults <- adults[(adults$site == "ALN"),]

ALN_adults_CH <- data.frame(individuals = c(unique(ALN_adults$ring)))

ALN_years <- sort(unique(ALN_adults$year))

for (year in ALN_years) {
  ALN_adults_CH[[as.character(year)]] <- NA
}

for (i in 1:nrow(ALN_adults_CH)) {
  individual <- ALN_adults_CH$individuals[i]
  capture_years <- unique(ALN_adults$year[ALN_adults$ring == individual])
  ALN_adults_CH[i, as.character(capture_years)] <- 1
}

ALN_adults_CH[is.na(ALN_adults_CH)] <- 0
ALN_adults_CH_matrix <- as.matrix(ALN_adults_CH[,-1])
ALN_adults_CH_matrix <- (apply(ALN_adults_CH_matrix, 2, as.numeric))

ALN_adults_CH_matrix

ALN_adults_marray <- marray(ALN_adults_CH_matrix)

ALN_adults_marray
```

The M-array above, generated from capture histories above the m-array using Kery and Schaub's marray function, shows individual capture histories at site 'ALN' collapsed into, on the diagonal, individuals marked and released on the nth occasion and recaptured on the n+1th occasion. The next column over on the same row shows individuals released on the nth occasion and recaptured on the n+2nd occasion. The far right column represents individuals marked and never recaptured. This is the format the IPM will take for the CMR portion of the model to estimate apparent survival. 

```{r creating m-array list for all sites, echo=FALSE}
create_marray <- function(site_code, data) {
  site_data <- data[data$site == site_code,]
  
  capture_history <- data.frame(individuals = unique(site_data$ring))
  site_years <- 2014:2025  
  #change the above line to sort(unique(site_data$year))
  for (year in site_years) {
    capture_history[[as.character(year)]] <- NA
  }
  
  for (i in 1:nrow(capture_history)) {
    individual <- capture_history$individuals[i]
    capture_years <- unique(site_data$year[site_data$ring == individual])
    capture_history[i, as.character(capture_years)] <- 1
  }
  
  capture_history[is.na(capture_history)] <- 0
  
  capture_history_matrix <- as.matrix(capture_history[,-1])
  capture_history_matrix <- apply(capture_history_matrix, 2, as.numeric)
  
  marray_result <- marray(capture_history_matrix)
  return(marray_result)
}

site_codes <- unique(adults$site)

marray_list <- lapply(site_codes, function(site) create_marray(site, adults))

```

##Plotting IPM inputs

The IPM takes population size data - in the form of maximum occupancy - as y_mat, the number of successful fledglings at each site each year - as J_mat, and the number of broods surveyed at each site in each year - as R_mat. These are plotted below:

```{r, echo=FALSE}

names(marray_list) <- site_codes

#max occupancy
phendat$ID <- as.factor(paste(phendat$year, phendat$site, phendat$box, sep = "_")) 
phendat$suc[phendat$suc == "-999"] <- 0
phendat <- phendat[!duplicated(phendat$ID, fromLast = TRUE), ]

ord_dates_phendat <- select(phendat, c('year', 'site', 'ID', 'n1', 'nl', 'latestfed', 'latestcc', 'fed', 'cc', 'cs', 'fki', 'hatching_first_recorded', 'v1date', 'v2date'))


transform_dates <- function(df) {
  date_columns <- 70:180
  occupancy_df <- matrix(0, ncol = length(date_columns), nrow = nrow(df))
  colnames(occupancy_df) <- as.character(date_columns)
  
  for (i in 1:nrow(df)) {
    dates <- suppressWarnings(as.numeric(df[i, c('n1', 'nl', 'latestfed', 'latestcc', 'fed', 'cc', 
                                                 'cs', 'fki', 'hatching_first_recorded', 'v1date', 
                                                 'v2date')]))    
    date_range <- dates[!is.na(dates)]
    if (length(date_range) > 0) {
      min_date <- max(min(date_range), 70)
      max_date <- min(max(date_range), 180)
      valid_dates <- min_date:max_date
      valid_dates <- valid_dates[valid_dates %in% date_columns]
      occupancy_df[i, as.character(valid_dates)] <- 1
    }
  }
  
  occupancy_df <- as.data.frame(occupancy_df)
  occupancy_df <- cbind(df[, c('year', 'site', 'ID')], occupancy_df)
  
  return(occupancy_df)
}

occupancy_data <- transform_dates(ord_dates_phendat)

max_occupancy <- function(df) {
  numeric_columns <- as.character(70:180)
  df[numeric_columns] <- lapply(df[numeric_columns], as.numeric)
  results <- df %>%
    group_by(year, site) %>%
    summarise(max_occupancy = max(colSums(across(all_of(numeric_columns)), na.rm = TRUE), na.rm = TRUE)) %>%
    ungroup()
  
  return(results)
}

max_occupancy_df <- max_occupancy(occupancy_data)

max_occupancy_df_lat <- max_occupancy_df %>%
  left_join(sitedat %>% select(site, Mean.Lat), by = "site")

occupancy_p <- ggplot(max_occupancy_df_lat, aes(x = factor(year), y = max_occupancy, fill = Mean.Lat, colour = Mean.Lat)) +
  scale_colour_gradient(high = 'purple3', low = 'orange') +
  scale_fill_gradient(high = 'purple3', low = 'orange') +
  #geom_violin(fill = "lightgray", alpha = 0.3, color = NA) +
  geom_jitter(shape = 21, size = 1, width = 0.2, height = 0) +
  geom_line(aes(group = site, colour = Mean.Lat)) +
  theme_minimal() +
  scale_y_continuous(n.breaks = 9) +
  labs(title = "Annual Occupancy by Site",
       x = "Year",
       y = "Occupancy",
       fill = "Mean Latitude",
       colour = 'Mean Latitude') +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_line(color = "gray", size = 0.5)
  )

#number of offspring

count_fledglings <- function(df) {
  results <- df %>%
    group_by(year, site) %>%
    summarise(fledgling_count = sum(suc, na.rm = TRUE)) %>%
    ungroup()
  
  return(results)
}

n_offspring_df <- count_fledglings(phendat)

n_offspring_df_lat <- n_offspring_df %>%
  left_join(sitedat %>% select(site, Mean.Lat), by = "site")

n_fledge_p <- ggplot(n_offspring_df_lat, aes(x = factor(year), y = fledgling_count, fill = Mean.Lat, colour = Mean.Lat)) +
  scale_colour_gradient(high = 'purple3', low = 'orange') +
  scale_fill_gradient(high = 'purple3', low = 'orange') +
  #geom_violin(fill = "lightgray", alpha = 0.3, color = NA) +
  geom_jitter(shape = 21, size = 1, width = 0.2, height = 0) +
  geom_line(aes(group = site, colour = Mean.Lat)) +
  theme_minimal() +
  labs(title = "Annual fledgling count by Site",
       x = "Year",
       y = "N fledglings",
       fill = "Mean Latitude",
       colour = "Mean Latitude") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_line(color = "gray", size = 0.5)
  )

count_broods_surveyed <- function(df) {
  numeric_columns <- as.character(70:180)
  df[numeric_columns] <- lapply(df[numeric_columns], as.numeric)
  results <- df %>%
    group_by(year, site) %>%
    summarise(brood_count = sum(rowSums(across(all_of(numeric_columns))) > 0, na.rm = TRUE)) %>%    ungroup()
  
  return(results)
}

broods_surveyed_df <- count_broods_surveyed(occupancy_data)

broods_surveyed_df_lat <- broods_surveyed_df %>%
  left_join(sitedat %>% select(site, Mean.Lat), by = "site")

broods_p <- ggplot(broods_surveyed_df_lat, aes(x = factor(year), y = brood_count, fill = Mean.Lat, colour = Mean.Lat)) +
  scale_colour_gradient(high = 'purple3', low = 'orange') +
  scale_fill_gradient(high = 'purple3', low = 'orange') +
  #geom_violin(fill = "lightgray", alpha = 0.3, color = NA) +
  geom_jitter(shape = 21, size = 1, width = 0.2, height = 0) +
  geom_line(aes(group = site, colour = Mean.Lat)) +
  theme_minimal() +
  labs(title = "Annual broods surveyed by Site",
       x = "Year",
       y = "N broods",
       fill = "Mean Latitude",
       colour = "Mean Latitude") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_line(color = "gray", size = 0.5)
  )

all_combinations <- expand.grid(site = site_codes, year = 2014:2025)

broods_surveyed_df <- all_combinations %>%
  left_join(broods_surveyed_df, by = c("site", "year")) %>%
  replace_na(list(brood_count = 0))

y_list <- c()
J_list <- c()
R_list <- c()

for (i in site_codes) {
  y_list[[i]] <- max_occupancy_df %>% filter(site == i) %>% arrange(year) %>% pull(max_occupancy)
  J_list[[i]] <- n_offspring_df   %>% filter(site == i) %>% arrange(year) %>% pull(fledgling_count)
  R_list[[i]] <- broods_surveyed_df %>% filter(site == i) %>% arrange(year) %>% pull(brood_count)
}


(occupancy_p / broods_p / n_fledge_p)

# Exercise 4: Once blue tit data are in format similar to here, run the Stan model on blue tit dataset and evaluate.

list_to_mat <- function(lst, site_codes, nyears) {
  mat <- matrix(0, nrow = length(site_codes), ncol = nyears,
                dimnames = list(site_codes, NULL))
  for (i in seq_along(site_codes)) {
    vals <- unlist(lst[[site_codes[i]]])
    mat[i, seq_along(vals)] <- vals
  }
  mat
}

nsites_bt <- 44
nyears_bt <- 12

y_mat_bt <- list_to_mat(y_list, site_codes, nyears_bt)
J_mat_bt <- list_to_mat(J_list, site_codes, nyears_bt)
R_mat_bt <- list_to_mat(R_list, site_codes, nyears_bt)


#Y - pop counts (max occ)
#J - nestling counts
#R - surveyed broods

```

##IPM data formatting

In the meantime, I'm working my way through the Rstan hoopoe IPM to see where I need to alter it for the blue tit data. I've removed all lines relating to phij as I have not looked at fledgling survival yet. I'm now trying to parse what each chunk is doing so that I can identify where nyears - 1 needs to be changed to nyears (the hoopoe data takes nyears-1 for # of fledglings and also # of broods surveyed, while we have data for these:

```{r Data Formatting, echo=TRUE}
#blue tit data format
print(dim(y_mat_bt))
print(dim(J_mat_bt))
print(dim(R_mat_bt))
print(dim(marray_list))
print("this returns NULL but is a list of 44 m-arrays containing up to 12 years of releases and 11 years of recaptures (n years by n-1 years)")

#Whereas for the hoopoes - which is the format taken by the RStan code
print(dim(y_mat))
print(dim(J_mat))
print(dim(R_mat))
print(dim(marray_j_3d))
print(dim(marray_a_3d))
```

As we can see above, J_mat and R_mat take dimensions (site, nyear-1) in the hoopoe model, whereas for the blue tit data they take (site, nyear). However, for the M-arrays, both Hoopoe and blue tit take (site, nyear-1, nyear). Therefore, some nyear-1s in the stan file need editing to nyear while others need leaving. 

Following Sanjana's email, I now understand that my data should be formatted the same as the Hoopoe data. This is because the nestling counts and surveyed broods do not exist at t=1, but somewhere between t=t and t=t+1. Therefore, the 12th entry should not exist from the model's perspective as the 12th year is still pre breeding - and from the model perspective, is until the 13th year. I have removed the 12th year below, as well as reformatting each input as.integer: 

```{r, echo=FALSE}
y_mat_bt    <- apply(y_mat_bt,    2, as.integer)
J_mat_bt <- apply(J_mat_bt[,1:11], 2, as.integer)
R_mat_bt <- apply(R_mat_bt[,1:11], 2, as.integer)
marray_list <- array(unlist(marray_list), dim = c(nsites_bt, nyears_bt - 1, nyears_bt))
marray_list <- apply(marray_list, c(1,2,3), as.integer)

print(dim(y_mat_bt))
print(dim(J_mat_bt))
print(dim(R_mat_bt))
print(dim(marray_list))
```

Next, I have repurposed the hoopoe code for the blue tits to fit the model. However, the following is causing issues:

"       N1 = lapply(1:nsites, function(s) round(runif(nyears,  1, 50))),
       NadSurv = lapply(1:nsites, function(s) round(runif(nyears,  5, 50))),
       Nadimm  = lapply(1:nsites, function(s) round(runif(nyears,  1, 50)))"

I think this is because some sites have very low occupancy, which is where we draw our Ntot estimate. This then pulls the estimate down towards 0, making it very easy for the number of adults surviving to the next census to exceed the number of adults. This is, naturally, impossible. To fix this maybe we add a parameter similar to recapture probability? That can vary site to site? That tells us that occupancy is in some way proportional to Ntot but not representative. 

I'm also worrying a bit about sites and boxes that weren't added until later years - I think they're being included in the model as empty rather than absent for the years before their implementation. This would create an artificial spike in lambda/ntot from 2016-2017 for example. I'm not sure how this would resolve as the model seems pretty inflexible on NAs and the dimensions of input data.

In meeting 18/06, we discussed somehow jittering the lines so that overlapping lines are clear. We also discussed altering the pallet to put sites in buckets rather than using a gradient. 
Possibility to tell the model to ignore -1s (to stand as NAs) - Sanjana has gotten around this by grouping the sites into 12 groups based on pairwise distances and elevation. It's probably better given the time constraint to stick with the grouping, but as a sensitivity analysis we could alter the resolution of the groupings (ie 22 groups) 
Life cycle diagram for methods - look at some write up on methods also 
Look at 

Meeting 25/06 
- revisit max occupancy and see if I get the same by counting entries with suc, fki, cc
- get IPM to work - fix marray, try to run sanjana's ipm on the unbucketed sites (it doesn't have n must be between 0 and N)
- think about priors (and priors sensitivity - what happens if we change our assumptions) - everything weak apart from juvenile survival
- continue to plot data input
- write methods - how the model uses the data 
- start to pull out estimates - iccs, lambdas for site years, plot with demographic rates 
- plot iccs and random effect variances (site-year) like hoopoes
- look into phij

simultaneous equation for ipm explanation

plot lambda (Y axis) (sitexyear) against demographic rates

(1,2) is year 1 site 2 in posterior df

scatter plot 

correlation = demographic rate driving 

plot the year variance with the ICCs 

model similar to initial sucmod for Ntot - ICC calc from this 

write up - finish methods, start results 