---
title: "Synchrony_Dissertation_Process"
author: "Ester Kirby"
date: "2026-06-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synchrony Dissertation Process

This is an RMarkdown serving as a data management plan and lab book for a dissertation undertaken in partial fulfillment of an MSc in Ecology, Evolution and Biodiversity (2026). 

``` {r, echo=FALSE}
library(lme4)
library(insight)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(viridis)
library(rstan)
```

```{r Importing and Cleaning Data}
#importing bird phenology dataframe
phendat <- read.csv("data/Bird_Phenology.csv")

#importing site details dataframe
sitedat <- read.csv("data/Site_Details.csv")

#importing adults dataframe
adults <- read.csv("data/Adults.csv")

#importing nestlings dataframe
nestlings <- read.csv("data/Nestlings.csv")

#changing -999s (0s in fledgling success caused by predation) to 0s
phendat$suc[phendat$suc == "-999"] <- 0

#checking for duplicates
phendat$ID <- as.factor(paste(phendat$year, phendat$site, phendat$box, sep = "_")) 
length(phendat$ID)
length(unique(phendat$ID))
phendat$ID[duplicated(phendat$ID)]
#2230 entries, 5 of which not unique

#keeping just the last brood (completed)
phendat <- phendat[!duplicated(phendat$ID, fromLast = TRUE), ]

phendat$ID[duplicated(phendat$ID)]
#success!

#removing non focal species
bludat <- phendat[phendat$species == "bluti",]
```

Above, I imported the relevant dataframes from the phenoweb project. 
I then changed -999s in column 'suc' to 0s in phendat as these indicated 0 success due to predation. This is because predation is impacted by and has impacts on synchrony, and therefore is biologically significant and does not need to be flagged as anomalous. The only concern is if the nestboxes cause a change in predation probability from natural levels - a literature search will be conducted to advise on this. 
I then identified duplicates, of which there were only 5. These were caused by birds abandoning nests and then new broods taking. I removed the first broods from the data as they did not represent proper brood attempts.
Finally, I created a separate dataframe containing only rows representing occupation by blue tits. 

```{r pressure, echo=FALSE}

#very early model attempt on fledgling success
bludat$site_year <- as.factor(paste(bludat$site, bludat$year, sep = "_")) 
sucmod <- lmer(suc ~ (1|year) + (1|site) + (1|site_year), data = bludat) 
summary(sucmod)

get_variance_intercept(sucmod)

#extracting the variance components 
var_components <- VarCorr(sucmod)
var_intercept_site_year <- attr(var_components$site_year, "stddev")^2
var_intercept_site <- attr(var_components$site, "stddev")^2
var_intercept_year <- attr(var_components$year, "stddev")^2

#a crude look at ICC synchrony from the model 
sucsync <- var_intercept_year/(var_intercept_year + var_intercept_site_year)

sucsync
```

##Hoopoe Integrated Population Model from Kery and Schaub 2012
#Translated to Rstan by Sanjana 

```{r, echo=FALSE}
# Simulate datasets for two more sites
nsites <- 3
nyears <- 9

# Note: CJS model is specified as a multinomial likelihood (i.e. individual capture histories are collapsed into m-arrays) and were not specified as individual-level state-space formulation I had done previously for blue tits
# More details on m-arrays and code to convert individual capture histories to m-arrays can be found in Chapter 7.10 in book, "Bayesian Population Analysis using WinBUGS --- A Hierarchical Perspective" (2012) by Marc Kéry and Michael Schaub.

# All site-1 associated data is the hoopoe dataset from Chapter 11, Kery & Schaub 2012 book
# Capture-recapture data: m-array of juveniles and adults (these are males and females together)
marray_j_site1 <- structure(
  c(15, 0, 0, 0, 0, 0, 0, 0, 3, 34, 0, 0, 0, 0, 0, 0,
    0, 9, 56, 0, 0, 0, 0, 0, 0, 1, 8, 48, 0, 0, 0, 0, 0, 0, 1, 3,
    45, 0, 0, 0, 0, 0, 0, 1, 13, 27, 0, 0, 0, 0, 0, 0, 2, 7, 37,
    0, 0, 0, 0, 0, 0, 0, 3, 39, 198, 287, 455, 518, 463, 493, 434, 405),
  .Dim = 8:9)

marray_a_site1 <- structure(
  c(14, 0, 0, 0, 0, 0, 0, 0, 2, 22, 0, 0, 0, 0, 0, 0,
    0, 4, 34, 0, 0, 0, 0, 0, 0, 0, 2, 51, 0, 0, 0, 0, 0, 0, 0, 3,
    45, 0, 0, 0, 0, 0, 0, 0, 3, 44, 0, 0, 0, 0, 0, 0, 0, 3, 48, 0,
    0, 0, 0, 0, 0, 0, 2, 51, 43, 44, 79, 94, 118, 113, 99, 90),
  .Dim = 8:9)

y_site1 <- c(32, 42, 64, 85, 82, 78, 73, 69, 79) # Population count data  (the maximal number of simultaneously occupied nest boxes in each year)
J_site1 <- c(189, 274, 398, 538, 520, 476, 463, 438) # Number of offspring
R_site1 <- c(28, 36, 57, 77, 81, 83, 77, 72) # Number of surveyed broods

# Site 2 
set.seed(42)
y_site2 <- round(y_site1 * 0.60 * exp(rnorm(nyears, 0, 0.10)))
J_site2 <- round(J_site1 * 0.60 * exp(rnorm(nyears - 1, 0, 0.10)))
R_site2 <- round(R_site1 * 0.60 * exp(rnorm(nyears - 1, 0, 0.10)))

view(y_site2)

marray_j_site2 <- structure(
  c(12, 0, 0, 0, 0, 0, 0, 0, 2, 28, 0, 0, 0, 0, 0, 0,
    0, 7, 45, 0, 0, 0, 0, 0, 0, 2, 6, 38, 0, 0, 0, 0, 0, 0, 2, 4,
    36, 0, 0, 0, 0, 0, 0, 1, 10, 22, 0, 0, 0, 0, 0, 0, 1, 5, 29,
    0, 0, 0, 0, 0, 0, 0, 2, 31, 210, 265, 430, 490, 440, 470, 410, 385),
  .Dim = 8:9)

marray_a_site2 <- structure(
  c(11, 0, 0, 0, 0, 0, 0, 0, 3, 19, 0, 0, 0, 0, 0, 0,
    0, 3, 29, 0, 0, 0, 0, 0, 0, 0, 3, 44, 0, 0, 0, 0, 0, 0, 0, 2,
    38, 0, 0, 0, 0, 0, 0, 0, 2, 38, 0, 0, 0, 0, 0, 0, 0, 2, 41, 0,
    0, 0, 0, 0, 0, 0, 1, 44, 40, 38, 68, 82, 105, 100, 88, 78),
  .Dim = 8:9)

# Site 3 
set.seed(43)
y_site3 <- round(y_site1 * 1.20 * exp(rnorm(nyears, 0, 0.12)))
J_site3 <- round(J_site1 * 1.20 * exp(rnorm(nyears - 1, 0, 0.12)))
R_site3 <- round(R_site1 * 1.20 * exp(rnorm(nyears - 1, 0, 0.12)))

marray_j_site3 <- structure(
  c(18, 0, 0, 0, 0, 0, 0, 0, 4, 40, 0, 0, 0, 0, 0, 0,
    0, 11, 63, 0, 0, 0, 0, 0, 0, 1, 10, 55, 0, 0, 0, 0, 0, 0, 1, 4,
    52, 0, 0, 0, 0, 0, 0, 2, 15, 31, 0, 0, 0, 0, 0, 0, 2, 9, 43,
    0, 0, 0, 0, 0, 0, 0, 4, 45, 185, 305, 478, 540, 480, 510, 452, 420),
  .Dim = 8:9)

marray_a_site3 <- structure(
  c(16, 0, 0, 0, 0, 0, 0, 0, 2, 25, 0, 0, 0, 0, 0, 0,
    0, 5, 39, 0, 0, 0, 0, 0, 0, 0, 2, 58, 0, 0, 0, 0, 0, 0, 0, 4,
    51, 0, 0, 0, 0, 0, 0, 0, 4, 50, 0, 0, 0, 0, 0, 0, 0, 4, 54, 0,
    0, 0, 0, 0, 0, 0, 3, 57, 46, 50, 88, 102, 128, 120, 108, 97),
  .Dim = 8:9)

# Combine data from all sites
# Combining pop counts
(y_mat <- rbind(y_site1, y_site2, y_site3))

# Combining offpsring numbers (J) and surveyed broods (R)
# J, R: [nsites, nyears-1] format
(J_mat <- rbind(J_site1, J_site2, J_site3))
(R_mat <- rbind(R_site1, R_site2, R_site3))

# Combining recapture histories (m-array format)
# marray: [nsites, nyears-1, nyears] format
marray_j_3d <- array(NA_integer_, dim = c(nsites, nyears - 1, nyears))
marray_a_3d <- array(NA_integer_, dim = c(nsites, nyears - 1, nyears))

marray_j_3d[1,,] <- marray_j_site1
marray_j_3d[2,,] <- marray_j_site2
marray_j_3d[3,,] <- marray_j_site3

marray_a_3d[1,,] <- marray_a_site1
marray_a_3d[2,,] <- marray_a_site2
marray_a_3d[3,,] <- marray_a_site3

# Put together all data to feed into stan model
stan_data <- list(
  nyears    = nyears,
  nsites    = nsites,
  y         = y_mat,      
  J         = J_mat,      
  R         = R_mat,        
  marray_j  = marray_j_3d, 
  marray_a  = marray_a_3d  
)

# Give Stan parameters to estimate
params <- c(
  "phij", "phia", "f", "omega", "p",
  "mphij", "mphia", "mfec", "mim",
  "lambda", "mlam",
  "sig_phij",      "sig_phia",      "sig_fec",      "sig_im",
  "sig_site_phij", "sig_site_phia", "sig_site_fec", "sig_site_im",
  "sig_sy_phij",   "sig_sy_phia",   "sig_sy_fec",   "sig_sy_im",
  "N1", "NadSurv", "Nadimm", "Ntot",
  "var_phij_year", "var_phij_siteyear", "icc_phij",
  "var_phia_year", "var_phia_siteyear", "icc_phia",
  "var_fec_year", "var_fec_siteyear", "icc_fec",
  "var_im_year", "var_im_siteyear", "icc_im"
)

# Setting stan model options
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
set.seed(123)

# MCMC settings
ni <- 20000   
nb <- 10000   
nt <- 5       
nc <- 4

## Initial values
inits <- lapply(1:nc, function(i) {
  list(l_mphij = rnorm(1, 0.2, 0.5),
       l_mphia = rnorm(1, 0.2, 0.5),
       l_mfec = rnorm(1, 0.2, 0.5),
       l_mim = rnorm(1, 0.2, 0.5),
       l_p = rnorm(1, 0.2, 1),
       sig_phij = runif(1, 0.1, 10),
       sig_phia = runif(1, 0.1, 10),
       sig_fec = runif(1, 0.1, 10),
       sig_im = runif(1, 0.1, 10),
       sig_site_phij = runif(1, 0.1, 10),
       sig_site_phia = runif(1, 0.1, 10),
       sig_site_fec = runif(1, 0.1, 10),
       sig_site_im = runif(1, 0.1, 10),
       sig_sy_phij = runif(1, 0.1, 10),
       sig_sy_phia = runif(1, 0.1, 10),
       sig_sy_fec = runif(1, 0.1, 10),
       sig_sy_im = runif(1, 0.1, 10),
       N1      = lapply(1:nsites, function(s) round(runif(nyears,  1, 50))),
       NadSurv = lapply(1:nsites, function(s) round(runif(nyears,  5, 50))),
       Nadimm  = lapply(1:nsites, function(s) round(runif(nyears,  1, 50)))
)})

# Run stan model
ipm_hoopoe <- stan(
  file    = "ipm_hoopoe.stan",
  data    = stan_data,
  init    = inits,
  pars    = params,
  chains  = nc,
  iter    = ni,
  warmup  = nb,
  thin    = nt,
  seed    = 2,
  control = list(
    adapt_delta   = 0.99,
    max_treedepth = 15    
  )
)

# Print stan model output
print(ipm_hoopoe,
      pars = c("mphij", "mphia", "mfec", "mim",
               "sig_phij",      "sig_phia",      "sig_fec",      "sig_im",
               "sig_site_phij", "sig_site_phia", "sig_site_fec", "sig_site_im",
               "sig_sy_phij",   "sig_sy_phia",   "sig_sy_fec",   "sig_sy_im",
               "mlam",   "var_phij_year", "var_phij_siteyear", "icc_phij",
               "var_phia_year", "var_phia_siteyear", "icc_phia",
               "var_fec_year", "var_fec_siteyear", "icc_fec",
               "var_im_year", "var_im_siteyear", "icc_im"),
      digits = 3)

# Get posterior distributions
posterior_ipm <- as.data.frame(ipm_hoopoe)

# extract summary (mean + 95% CrI) for parameters
extract_summary <- function(posterior, pattern) {
  cols <- grep(pattern, names(posterior), value = TRUE)
  data.frame(
    param = cols,
    mean  = colMeans(posterior[, cols, drop = FALSE]),
    lower = apply(posterior[, cols, drop = FALSE], 2, quantile, 0.025),
    upper = apply(posterior[, cols, drop = FALSE], 2, quantile, 0.975)
  )
}

# Get nyears and nsites
nyears  <- stan_data$nyears
nsites  <- stan_data$nsites
years   <- 1:nyears
years_m1 <- 1:(nyears - 1)  # for parameters indexed over nyears-1

site_labels <- paste0("Site ", 1:nsites)

# Estimated Ntot per site per year
ntot_list <- lapply(1:nsites, function(s) {
  pattern <- paste0("Ntot\\[", s, ",")
  df <- extract_summary(posterior_ipm, pattern)
  df$year <- years
  df$site <- site_labels[s]
  df
})
ntot_df <- bind_rows(ntot_list)

# Observed counts per site per year
obs_counts <- expand.grid(site = site_labels, year = years) %>%
  arrange(site, year) %>%
  mutate(observed = as.vector(t(stan_data$y)))  # y is [nsites, nyears]

# Merge both dfs
ntot_df <- left_join(ntot_df, obs_counts, by = c("site", "year"))

# Plot
p_ntot <- ggplot(ntot_df, aes(x = year)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = mean, colour = "Estimated Ntot"), linewidth = 0.9) +
  geom_point(aes(y = mean, colour = "Estimated Ntot"), size = 1.5) +
  geom_line(aes(y = observed, colour = "Observed count"),
            linewidth = 0.9, linetype = "dashed") +
  geom_point(aes(y = observed, colour = "Observed count"), size = 1.5) +
  scale_colour_manual(values = c("Estimated Ntot" = "steelblue",
                                 "Observed count"  = "firebrick")) +
  facet_wrap(~ site, scales = "free_y") +
  labs(title = "Population size", x = "Year", y = "N",
       colour = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom") +
  scale_x_continuous(n.breaks=8)

p_ntot

# Exercises for Ester -

# Exercise1: Plot and investigate temporal trends in other demographic parameters (survival, productivity, immigration, population growth rate etc) across the different sites (doing this will be handy for blue tit data analysis)

#productivity/fecundity 
# Estimated f per site per year
f_list <- lapply(1:nsites, function(s) {
  pattern <- paste0("f\\[", s, ",")
  df <- extract_summary(posterior_ipm, pattern)
  df$year <- years_m1
  df$site <- site_labels[s]
  df
})
f_est <- bind_rows(f_list)

# Observed f per site per year
obs_f <- expand.grid(site = site_labels, year = years_m1) %>%
  arrange(site, year) %>%
  mutate(observed = as.vector(t(stan_data$J / stan_data$R)))

# Merge both dfs
f_df <- left_join(f_est, obs_f, by = c("site", "year"))

# Plot
p_f <- ggplot(f_df, aes(x = year)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = mean, colour = "Estimated f"), linewidth = 0.9) +
  geom_point(aes(y = mean, colour = "Estimated f"), size = 1.5) +
  geom_line(aes(y = observed, colour = "Observed f"),
            linewidth = 0.9, linetype = "dashed") +
  geom_point(aes(y = observed, colour = "Observed f"), size = 1.5) +
  scale_colour_manual(values = c("Estimated f" = "steelblue",
                                 "Observed f"  = "firebrick")) +
  facet_wrap(~ site, scales = "free_y") +
  labs(title = "Fecundity per Nest", x = "Year", y = "N Fledglings/female",
       colour = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom") +
  scale_x_continuous(n.breaks=8)

p_f
```


