---
title: "Synchrony_Dissertation_Process"
author: "Ester Kirby"
date: "2026-06-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synchrony Dissertation Process

This is an RMarkdown serving as a data management plan and lab book for a dissertation undertaken in partial fulfillment of an MSc in Ecology, Evolution and Biodiversity (2026). 

``` {r, echo=FALSE}
library(lme4)
library(insight)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(viridis)
library(rstan)
library(patchwork)
```

```{r Importing and Cleaning Data}
#importing bird phenology dataframe
phendat <- read.csv("data/Bird_Phenology.csv")

#importing site details dataframe
sitedat <- read.csv("data/Site_Details.csv")

#importing adults dataframe
adults <- read.csv("data/Adults.csv")
adults <- adults[adults$season == 'spring',]

#importing nestlings dataframe
nestlings <- read.csv("data/Nestlings.csv")

#changing -999s (0s in fledgling success caused by predation) to 0s
phendat$suc[phendat$suc == "-999"] <- 0

#checking for duplicates
phendat$ID <- as.factor(paste(phendat$year, phendat$site, phendat$box, sep = "_")) 
length(phendat$ID)
length(unique(phendat$ID))
phendat$ID[duplicated(phendat$ID)]
#2230 entries, 5 of which not unique

#keeping just the last brood (completed)
phendat <- phendat[!duplicated(phendat$ID, fromLast = TRUE), ]

phendat$ID[duplicated(phendat$ID)]
#success!

#removing non focal species
bludat <- phendat[phendat$species == "bluti",]
```

Above, I imported the relevant dataframes from the phenoweb project. 
I then changed -999s in column 'suc' to 0s in phendat as these indicated 0 success due to predation. This is because predation is impacted by and has impacts on synchrony, and therefore is biologically significant and does not need to be flagged as anomalous. The only concern is if the nestboxes cause a change in predation probability from natural levels - a literature search will be conducted to advise on this. 
I then identified duplicates, of which there were only 5. These were caused by birds abandoning nests and then new broods taking. I removed the first broods from the data as they did not represent proper brood attempts.
Finally, I created a separate dataframe containing only rows representing occupation by blue tits. 

```{r clutch swap treatment}
bludatswap <- bludat %>%
  mutate(clutch.swap.treatment = ifelse(clutch.swap.treatment %in% 1:9, "swap", clutch.swap.treatment))

bludatswap <- bludatswap[bludatswap$year %in% c("2017", "2018", "2019"), ]

bludatswap$clutch.swap.treatment <- substr(bludatswap$clutch.swap.treatment, 1, 3)

clutchswapmod <- lm(suc ~ clutch.swap.treatment, data = bludatswap)
summary(clutchswapmod)

swapmodmixed <- lmer(suc ~ clutch.swap.treatment + (1|site) + (1|year), data = bludatswap)
summary(swapmodmixed)

#relevel to make unm the reference
str(bludatswap)
bludatswap$clutch.swap.treatment <- as.factor(bludatswap$clutch.swap.treatment)
bludatswap$clutch.swap.treatment <- relevel(bludatswap$clutch.swap.treatment, ref = "unm")

#making year a factor to reduce noise
bludatswap$year <- as.factor(bludatswap$year)

#removing nas in swap
bludatswap <- bludatswap %>%
  mutate(clutch.swap.treatment = na_if(clutch.swap.treatment, "")) %>%
  drop_na(clutch.swap.treatment)

#popping odd and not possible together 
bludatswap <- bludat %>%
  mutate(clutch.swap.treatment = as.character(clutch.swap.treatment)) %>%
  mutate(clutch.swap.treatment = ifelse(clutch.swap.treatment == "not", "odd", clutch.swap.treatment))

swapmodmixed <- lmer(suc ~ (1|clutch.swap.treatment) + year + (1|site), data = bludatswap)
summary(swapmodmixed)

tapply(bludatswap$clutch.swap.treatment, bludatswap$clutch.swap.treatment, length)
tapply(bludatswap$suc, bludatswap$clutch.swap.treatment, mean, na.rm=TRUE)

```

Above, I isolated the clutch swap experiment years and ran a linear mixed model to look for variance attributed to clutch swap treatment as opposed to unmanipulated clutches. This allowed me to decide whether to include these years, just the unmanipulated broods, or exclude entirely. Overall variance attributed to treatment was low compared to residual, and mean success was similar between unmanipulated and swapped clutches. Therefore, it appears the effect of the clutch swap experiment on fledgling success was negligible and these years can be included in full. 

```{r fledgling success initial glmm}

#very early model attempt on fledgling success
bludat$site_year <- as.factor(paste(bludat$site, bludat$year, sep = "_")) 
sucmod <- lmer(suc ~ (1|year) + (1|site) + (1|site_year), data = bludat) 
summary(sucmod)

get_variance_intercept(sucmod)

#extracting the variance components 
var_components <- VarCorr(sucmod)
var_intercept_site_year <- attr(var_components$site_year, "stddev")^2
var_intercept_site <- attr(var_components$site, "stddev")^2
var_intercept_year <- attr(var_components$year, "stddev")^2

#a crude look at ICC synchrony from the model 
sucsync <- var_intercept_year/(var_intercept_year + var_intercept_site_year)

sucsync
```

Here, I ran a simple linear mixed model on fledgling success with year, site and site year random effects. I then extracted the variance components and calculated the ICC of fledgling success synchrony

##Hoopoe Integrated Population Model from Kery and Schaub 2012
#Translated to Rstan by Sanjana 

```{r Sanjanas code, echo=FALSE}
# Simulate datasets for two more sites
nsites <- 3
nyears <- 9

# Note: CJS model is specified as a multinomial likelihood (i.e. individual capture histories are collapsed into m-arrays) and were not specified as individual-level state-space formulation I had done previously for blue tits
# More details on m-arrays and code to convert individual capture histories to m-arrays can be found in Chapter 7.10 in book, "Bayesian Population Analysis using WinBUGS --- A Hierarchical Perspective" (2012) by Marc Kéry and Michael Schaub.

# All site-1 associated data is the hoopoe dataset from Chapter 11, Kery & Schaub 2012 book
# Capture-recapture data: m-array of juveniles and adults (these are males and females together)
marray_j_site1 <- structure(
  c(15, 0, 0, 0, 0, 0, 0, 0, 3, 34, 0, 0, 0, 0, 0, 0,
    0, 9, 56, 0, 0, 0, 0, 0, 0, 1, 8, 48, 0, 0, 0, 0, 0, 0, 1, 3,
    45, 0, 0, 0, 0, 0, 0, 1, 13, 27, 0, 0, 0, 0, 0, 0, 2, 7, 37,
    0, 0, 0, 0, 0, 0, 0, 3, 39, 198, 287, 455, 518, 463, 493, 434, 405),
  .Dim = 8:9)

marray_a_site1 <- structure(
  c(14, 0, 0, 0, 0, 0, 0, 0, 2, 22, 0, 0, 0, 0, 0, 0,
    0, 4, 34, 0, 0, 0, 0, 0, 0, 0, 2, 51, 0, 0, 0, 0, 0, 0, 0, 3,
    45, 0, 0, 0, 0, 0, 0, 0, 3, 44, 0, 0, 0, 0, 0, 0, 0, 3, 48, 0,
    0, 0, 0, 0, 0, 0, 2, 51, 43, 44, 79, 94, 118, 113, 99, 90),
  .Dim = 8:9)

y_site1 <- c(32, 42, 64, 85, 82, 78, 73, 69, 79) # Population count data  (the maximal number of simultaneously occupied nest boxes in each year)
J_site1 <- c(189, 274, 398, 538, 520, 476, 463, 438) # Number of offspring
R_site1 <- c(28, 36, 57, 77, 81, 83, 77, 72) # Number of surveyed broods

# Site 2 
set.seed(42)
y_site2 <- round(y_site1 * 0.60 * exp(rnorm(nyears, 0, 0.10)))
J_site2 <- round(J_site1 * 0.60 * exp(rnorm(nyears - 1, 0, 0.10)))
R_site2 <- round(R_site1 * 0.60 * exp(rnorm(nyears - 1, 0, 0.10)))

marray_j_site2 <- structure(
  c(12, 0, 0, 0, 0, 0, 0, 0, 2, 28, 0, 0, 0, 0, 0, 0,
    0, 7, 45, 0, 0, 0, 0, 0, 0, 2, 6, 38, 0, 0, 0, 0, 0, 0, 2, 4,
    36, 0, 0, 0, 0, 0, 0, 1, 10, 22, 0, 0, 0, 0, 0, 0, 1, 5, 29,
    0, 0, 0, 0, 0, 0, 0, 2, 31, 210, 265, 430, 490, 440, 470, 410, 385),
  .Dim = 8:9)

marray_a_site2 <- structure(
  c(11, 0, 0, 0, 0, 0, 0, 0, 3, 19, 0, 0, 0, 0, 0, 0,
    0, 3, 29, 0, 0, 0, 0, 0, 0, 0, 3, 44, 0, 0, 0, 0, 0, 0, 0, 2,
    38, 0, 0, 0, 0, 0, 0, 0, 2, 38, 0, 0, 0, 0, 0, 0, 0, 2, 41, 0,
    0, 0, 0, 0, 0, 0, 1, 44, 40, 38, 68, 82, 105, 100, 88, 78),
  .Dim = 8:9)

# Site 3 
set.seed(43)
y_site3 <- round(y_site1 * 1.20 * exp(rnorm(nyears, 0, 0.12)))
J_site3 <- round(J_site1 * 1.20 * exp(rnorm(nyears - 1, 0, 0.12)))
R_site3 <- round(R_site1 * 1.20 * exp(rnorm(nyears - 1, 0, 0.12)))

marray_j_site3 <- structure(
  c(18, 0, 0, 0, 0, 0, 0, 0, 4, 40, 0, 0, 0, 0, 0, 0,
    0, 11, 63, 0, 0, 0, 0, 0, 0, 1, 10, 55, 0, 0, 0, 0, 0, 0, 1, 4,
    52, 0, 0, 0, 0, 0, 0, 2, 15, 31, 0, 0, 0, 0, 0, 0, 2, 9, 43,
    0, 0, 0, 0, 0, 0, 0, 4, 45, 185, 305, 478, 540, 480, 510, 452, 420),
  .Dim = 8:9)

marray_a_site3 <- structure(
  c(16, 0, 0, 0, 0, 0, 0, 0, 2, 25, 0, 0, 0, 0, 0, 0,
    0, 5, 39, 0, 0, 0, 0, 0, 0, 0, 2, 58, 0, 0, 0, 0, 0, 0, 0, 4,
    51, 0, 0, 0, 0, 0, 0, 0, 4, 50, 0, 0, 0, 0, 0, 0, 0, 4, 54, 0,
    0, 0, 0, 0, 0, 0, 3, 57, 46, 50, 88, 102, 128, 120, 108, 97),
  .Dim = 8:9)

# Combine data from all sites
# Combining pop counts
(y_mat <- rbind(y_site1, y_site2, y_site3))

# Combining offpsring numbers (J) and surveyed broods (R)
# J, R: [nsites, nyears-1] format
(J_mat <- rbind(J_site1, J_site2, J_site3))
(R_mat <- rbind(R_site1, R_site2, R_site3))

# Combining recapture histories (m-array format)
# marray: [nsites, nyears-1, nyears] format
marray_j_3d <- array(NA_integer_, dim = c(nsites, nyears - 1, nyears))
marray_a_3d <- array(NA_integer_, dim = c(nsites, nyears - 1, nyears))

marray_j_3d[1,,] <- marray_j_site1
marray_j_3d[2,,] <- marray_j_site2
marray_j_3d[3,,] <- marray_j_site3

marray_a_3d[1,,] <- marray_a_site1
marray_a_3d[2,,] <- marray_a_site2
marray_a_3d[3,,] <- marray_a_site3

# Put together all data to feed into stan model
stan_data <- list(
  nyears    = nyears,
  nsites    = nsites,
  y         = y_mat,      
  J         = J_mat,      
  R         = R_mat,        
  marray_j  = marray_j_3d, 
  marray_a  = marray_a_3d  
)

# Give Stan parameters to estimate
params <- c(
  "phij", "phia", "f", "omega", "p",
  "mphij", "mphia", "mfec", "mim",
  "lambda", "mlam",
  "sig_phij",      "sig_phia",      "sig_fec",      "sig_im",
  "sig_site_phij", "sig_site_phia", "sig_site_fec", "sig_site_im",
  "sig_sy_phij",   "sig_sy_phia",   "sig_sy_fec",   "sig_sy_im",
  "N1", "NadSurv", "Nadimm", "Ntot",
  "var_phij_year", "var_phij_siteyear", "icc_phij",
  "var_phia_year", "var_phia_siteyear", "icc_phia",
  "var_fec_year", "var_fec_siteyear", "icc_fec",
  "var_im_year", "var_im_siteyear", "icc_im"
)

# Setting stan model options
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
set.seed(123)

# MCMC settings
ni <- 20000   
nb <- 10000   
nt <- 5       
nc <- 4

## Initial values
inits <- lapply(1:nc, function(i) {
  list(l_mphij = rnorm(1, 0.2, 0.5),
       l_mphia = rnorm(1, 0.2, 0.5),
       l_mfec = rnorm(1, 0.2, 0.5),
       l_mim = rnorm(1, 0.2, 0.5),
       l_p = rnorm(1, 0.2, 1),
       sig_phij = runif(1, 0.1, 10),
       sig_phia = runif(1, 0.1, 10),
       sig_fec = runif(1, 0.1, 10),
       sig_im = runif(1, 0.1, 10),
       sig_site_phij = runif(1, 0.1, 10),
       sig_site_phia = runif(1, 0.1, 10),
       sig_site_fec = runif(1, 0.1, 10),
       sig_site_im = runif(1, 0.1, 10),
       sig_sy_phij = runif(1, 0.1, 10),
       sig_sy_phia = runif(1, 0.1, 10),
       sig_sy_fec = runif(1, 0.1, 10),
       sig_sy_im = runif(1, 0.1, 10),
       N1      = lapply(1:nsites, function(s) round(runif(nyears,  1, 50))),
       NadSurv = lapply(1:nsites, function(s) round(runif(nyears,  5, 50))),
       Nadimm  = lapply(1:nsites, function(s) round(runif(nyears,  1, 50)))
)})

# Run stan model
#ipm_hoopoe <- stan(
#  file    = "ipm_hoopoe.stan",
#  data    = stan_data,
#  init    = inits,
#  pars    = params,
#  chains  = nc,
#  iter    = ni,
#  warmup  = nb,
#  thin    = nt,
#  seed    = 2,
#  control = list(
#    adapt_delta   = 0.99,
#    max_treedepth = 15    
#  )
#)

#rds_file_path <- "ipm_hoopoe.rds"

#saveRDS(ipm_hoopoe, rds_file_path)

ipm_hoopoe <- readRDS("ipm_hoopoe.rds")

# Print stan model output
print(ipm_hoopoe,
      pars = c("mphij", "mphia", "mfec", "mim",
               "sig_phij",      "sig_phia",      "sig_fec",      "sig_im",
               "sig_site_phij", "sig_site_phia", "sig_site_fec", "sig_site_im",
               "sig_sy_phij",   "sig_sy_phia",   "sig_sy_fec",   "sig_sy_im",
               "mlam",   "var_phij_year", "var_phij_siteyear", "icc_phij",
               "var_phia_year", "var_phia_siteyear", "icc_phia",
               "var_fec_year", "var_fec_siteyear", "icc_fec",
               "var_im_year", "var_im_siteyear", "icc_im"),
      digits = 3)

# Get posterior distributions
posterior_ipm <- as.data.frame(ipm_hoopoe)

# extract summary (mean + 95% CrI) for parameters
extract_summary <- function(posterior, pattern) {
  cols <- grep(pattern, names(posterior), value = TRUE)
  data.frame(
    param = cols,
    mean  = colMeans(posterior[, cols, drop = FALSE]),
    lower = apply(posterior[, cols, drop = FALSE], 2, quantile, 0.025),
    upper = apply(posterior[, cols, drop = FALSE], 2, quantile, 0.975)
  )
}

# Get nyears and nsites
nyears  <- stan_data$nyears
nsites  <- stan_data$nsites
years   <- 1:nyears
years_m1 <- 1:(nyears - 1)  # for parameters indexed over nyears-1

site_labels <- paste0("Site ", 1:nsites)

# Estimated Ntot per site per year
ntot_list <- lapply(1:nsites, function(s) {
  pattern <- paste0("Ntot\\[", s, ",")
  df <- extract_summary(posterior_ipm, pattern)
  df$year <- years
  df$site <- site_labels[s]
  df
})
ntot_df <- bind_rows(ntot_list)

# Observed counts per site per year
obs_counts <- expand.grid(site = site_labels, year = years) %>%
  arrange(site, year) %>%
  mutate(observed = as.vector(t(stan_data$y)))  # y is [nsites, nyears]

# Merge both dfs
ntot_df <- left_join(ntot_df, obs_counts, by = c("site", "year"))

# Plot
p_ntot <- ggplot(ntot_df, aes(x = year)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = mean, colour = "Estimated Ntot"), linewidth = 0.9) +
  geom_point(aes(y = mean, colour = "Estimated Ntot"), size = 1.5) +
  geom_line(aes(y = observed, colour = "Observed count"),
            linewidth = 0.9, linetype = "dashed") +
  geom_point(aes(y = observed, colour = "Observed count"), size = 1.5) +
  scale_colour_manual(values = c("Estimated Ntot" = "steelblue",
                                 "Observed count"  = "firebrick")) +
  facet_wrap(~ site, scales = "free_y") +
  labs(title = "Population size", x = "Year", y = "N",
       colour = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom") +
  scale_x_continuous(n.breaks=8)

p_ntot

# Exercises for Ester -

# Exercise1: Plot and investigate temporal trends in other demographic parameters (survival, productivity, immigration, population growth rate etc) across the different sites (doing this will be handy for blue tit data analysis)

#productivity/fecundity 
# Estimated f per site per year
f_list <- lapply(1:nsites, function(s) {
  pattern <- paste0("f\\[", s, ",")
  df <- extract_summary(posterior_ipm, pattern)
  df$year <- years_m1
  df$site <- site_labels[s]
  df
})
f_est <- bind_rows(f_list)

# Observed f per site per year
obs_f <- expand.grid(site = site_labels, year = years_m1) %>%
  arrange(site, year) %>%
  mutate(observed = as.vector(t(stan_data$J / stan_data$R)))

# Merge both dfs
f_df <- left_join(f_est, obs_f, by = c("site", "year"))

# Plot
p_f <- ggplot(f_df, aes(x = year)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = mean, colour = "Estimated f"), linewidth = 0.9) +
  geom_point(aes(y = mean, colour = "Estimated f"), size = 1.5) +
  geom_line(aes(y = observed, colour = "Observed f"),
            linewidth = 0.9, linetype = "dashed") +
  geom_point(aes(y = observed, colour = "Observed f"), size = 1.5) +
  scale_colour_manual(values = c("Estimated f" = "steelblue",
                                 "Observed f"  = "firebrick")) +
  facet_wrap(~ site, scales = "free_y") +
  labs(title = "Fecundity per Nest", x = "Year", y = "N Fledglings/female",
       colour = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom") +
  scale_x_continuous(n.breaks=8)

p_f
```

```{r hoopoe plotting - Sanjanas exercises}
# Exercise1: Plot and investigate temporal trends in other demographic parameters (survival, productivity, immigration, population growth rate etc) across the different sites (doing this will be handy for blue tit data analysis)

#productivity/fecundity 
# Estimated f per site per year
f_list <- lapply(1:nsites, function(s) {
  pattern <- paste0("f\\[", s, ",")
  df <- extract_summary(posterior_ipm, pattern)
  df$year <- years_m1
  df$site <- site_labels[s]
  df
})
f_est <- bind_rows(f_list)

# Observed f per site per year
obs_f <- expand.grid(site = site_labels, year = years_m1) %>%
  arrange(site, year) %>%
  mutate(observed = as.vector(t(stan_data$J / stan_data$R)))

# Merge both dfs
f_df <- left_join(f_est, obs_f, by = c("site", "year"))

# Plot
p_f <- ggplot(f_df, aes(x = year)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = mean, colour = "Estimated f"), linewidth = 0.9) +
  geom_point(aes(y = mean, colour = "Estimated f"), size = 1.5) +
  geom_line(aes(y = observed, colour = "Observed f"),
            linewidth = 0.9, linetype = "dashed") +
  geom_point(aes(y = observed, colour = "Observed f"), size = 1.5) +
  scale_colour_manual(values = c("Estimated f" = "steelblue",
                                 "Observed f"  = "firebrick")) +
  facet_wrap(~ site, scales = "free_y") +
  labs(title = "Fecundity per Nest", x = "Year", y = "N Fledglings/female",
       colour = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom") +
  scale_x_continuous(n.breaks=8)

#p_f


#survival
#juveniles
# Estimated phij per site per year
phij_list <- lapply(1:nsites, function(s) {
  pattern <- paste0("phij\\[", s, ",")
  df <- extract_summary(posterior_ipm, pattern)
  df$year <- years_m1
  df$site <- site_labels[s]
  df
})
phij_est <- bind_rows(phij_list)

#adults
phia_list <- lapply(1:nsites, function(s) {
  pattern <- paste0("phia\\[", s, ",")
  df <- extract_summary(posterior_ipm, pattern)
  df$year <- years_m1
  df$site <- site_labels[s]
  df
})
phia_est <- bind_rows(phia_list)


# Observed f per site per year
#obs_phij <- expand.grid(site = site_labels, year = years_m1) %>%
 # arrange(site, year) %>%
  #mutate(observed = as.vector(t()))

# Merge both dfs
phi_df <- left_join(phij_est, phia_est, by = c("site", "year"))

# Plot
p_phi <- ggplot(phi_df, aes(x = year)) +
  geom_ribbon(aes(ymin = lower.x, ymax = upper.x), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = mean.x, colour = "Estimated juvenile survival"), linewidth = 0.9) +
  geom_point(aes(y = mean.x, colour = "Estimated juvenile survival"), size = 1.5) +
  geom_ribbon(aes(ymin = lower.y, ymax = upper.y), fill = "firebrick", alpha = 0.25) +
  geom_line(aes(y = mean.y, colour = "Estimated adult survival"), linewidth = 0.9) +
  geom_point(aes(y = mean.y, colour = "Estimated adult survival"), size = 1.5) +
  scale_colour_manual(values = c("Estimated juvenile survival" = "steelblue", "Estimated adult survival" = "firebrick")) +
  facet_wrap(~ site, scales = "free_y") +
  labs(title = "Estimated Survival", x = "Year", y = "Phi",
       colour = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom") +
  scale_x_continuous(n.breaks=8)

#p_phi

#population growth rate

lambda_list <- lapply(1:nsites, function(s) {
  pattern <- paste0("lambda\\[", s, ",")
  df <- extract_summary(posterior_ipm, pattern)
  df$year <- years_m1
  df$site <- site_labels[s]
  df
})
lambda_df <- bind_rows(lambda_list)

# Observed counts per site per year
#obs_counts <- expand.grid(site = site_labels, year = years) %>%
 # arrange(site, year) %>%
  #mutate(observed = as.vector(t(stan_data$y)))  # y is [nsites, nyears]

# Merge both dfs
#lambda_df <- left_join(lambda_df, obs_counts, by = c("site", "year"))

# Plot
p_lambda <- ggplot(lambda_df, aes(x = year)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = mean, colour = "Estimated lambda"), linewidth = 0.9) +
  geom_point(aes(y = mean, colour = "Estimated lambda"), size = 1.5) +
  #geom_line(aes(y = observed, colour = "Observed count"),
   #         linewidth = 0.9, linetype = "dashed") +
  #geom_point(aes(y = observed, colour = "Observed count"), size = 1.5) +
  scale_colour_manual(values = c("Estimated lambda" = "steelblue")) +
  facet_wrap(~ site, scales = "free_y") +
  labs(title = "Population growth rate", x = "Year", y = "Lambda",
       colour = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom") +
  scale_x_continuous(n.breaks=8)

#p_lambda

#immigration
Nadimm_list <- lapply(1:nsites, function(s) {
  pattern <- paste0("Nadimm\\[", s, ",")
  df <- extract_summary(posterior_ipm, pattern)
  df$year <- years
  df$site <- site_labels[s]
  df
})
Nadimm_df <- bind_rows(Nadimm_list)

# Observed counts per site per year
obs_counts <- expand.grid(site = site_labels, year = years) %>%
  arrange(site, year) %>%
  mutate(observed = as.vector(t(stan_data$y)))  # y is [nsites, nyears]

# Merge both dfs
Nadimm_df <- left_join(Nadimm_df, obs_counts, by = c("site", "year"))

# Plot
p_Nadimm <- ggplot(Nadimm_df, aes(x = year)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = mean, colour = "Estimated Nadimm"), linewidth = 0.9) +
  geom_point(aes(y = mean, colour = "Estimated Nadimm"), size = 1.5) +
#  geom_line(aes(y = observed, colour = "Observed count"),
#            linewidth = 0.9, linetype = "dashed") +
#  geom_point(aes(y = observed, colour = "Observed count"), size = 1.5) +
  scale_colour_manual(values = c("Estimated Nadimm" = "steelblue")) +
  facet_wrap(~ site, scales = "free_y") +
  labs(title = "N Immigrants", x = "Year", y = "N",
       colour = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom") +
  scale_x_continuous(n.breaks=8)

#p_Nadimm

rates_plots <- (p_ntot | p_lambda | p_f) / (p_phi | p_Nadimm)

rates_plots

# Exercise2: Plot and investigate posterior distributions of random effect variances and spatial synchrony (ICCs) for all demographic parameters (doing this will be handy for blue tit data analysis)

#lambda
lambda_pds <- posterior_ipm %>%
  select(contains("lambda")) %>%
  pivot_longer(cols = everything(), values_to = "Value")

pds_lambda <- ggplot(lambda_pds, aes(x = Value)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Pooled Posterior Distribution of Lambda",
       x = "Lambda Value",
       y = "Density") +
  theme_minimal()

#phia icc

phia_icc <- posterior_ipm %>%
  select(contains("icc_phia")) %>%
  pivot_longer(cols = everything(), values_to = "Value")

icc_phia_p <- ggplot(phia_icc, aes(x = Value)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Adult Survival ICC",
       x = "phia ICC",
       y = "Density") +
  theme_minimal()

#icc_phia_p

#var_phia_siteyear

var_phia_siteyear <- posterior_ipm %>%
  select(contains("var_phia_siteyear")) %>%
  pivot_longer(cols = everything(), values_to = "Value")

var_phia_siteyear_p <- ggplot(var_phia_siteyear, aes(x = Value)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Adult Survival Random Effect Variances",
       x = "phia var",
       y = "Density") +
  theme_minimal()

#var_phia_siteyear_p

#phij icc

phij_icc <- posterior_ipm %>%
  select(contains("icc_phij")) %>%
  pivot_longer(cols = everything(), values_to = "Value")

icc_phij_p <- ggplot(phij_icc, aes(x = Value)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Juvenile Survival ICC",
       x = "phij ICC",
       y = "Density") +
  theme_minimal()

#icc_phij_p

#var_phij_siteyear

var_phij_siteyear <- posterior_ipm %>%
  select(contains("var_phij_siteyear")) %>%
  pivot_longer(cols = everything(), values_to = "Value")

var_phij_siteyear_p <- ggplot(var_phij_siteyear, aes(x = Value)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Juvenile Survival Random Effect Variances",
       x = "phij var",
       y = "Density") +
  theme_minimal()

#var_phij_siteyear_p

#fec icc

fec_icc <- posterior_ipm %>%
  select(contains("icc_fec")) %>%
  pivot_longer(cols = everything(), values_to = "Value")

icc_fec_p <- ggplot(fec_icc, aes(x = Value)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Fecundity ICC",
       x = "fec ICC",
       y = "Density") +
  theme_minimal()

#icc_fec_p

#var_fec_siteyear

var_fec_siteyear <- posterior_ipm %>%
  select(contains("var_fec_siteyear")) %>%
  pivot_longer(cols = everything(), values_to = "Value")

var_fec_siteyear_p <- ggplot(var_fec_siteyear, aes(x = Value)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Fecundity Random Effect Variances",
       x = "fec var",
       y = "Density") +
  theme_minimal()

#var_fec_siteyear_p

#im icc

im_icc <- posterior_ipm %>%
  select(contains("icc_im")) %>%
  pivot_longer(cols = everything(), values_to = "Value")

icc_im_p <- ggplot(im_icc, aes(x = Value)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Immigration ICC",
       x = "im ICC",
       y = "Density") +
  theme_minimal()

#icc_im_p

#var_im_siteyear

var_im_siteyear <- posterior_ipm %>%
  select(contains("var_im_siteyear")) %>%
  pivot_longer(cols = everything(), values_to = "Value")

var_im_siteyear_p <- ggplot(var_im_siteyear, aes(x = Value)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Immigration Random Effect Variances",
       x = "im var",
       y = "Density") +
  theme_minimal()

#var_im_siteyear_p

icc_var_plots <- (var_phia_siteyear_p | icc_phia_p) / (var_phij_siteyear_p | icc_phij_p) / (var_fec_siteyear_p | icc_fec_p) / (var_im_siteyear_p | icc_im_p)

icc_var_plots
```

``` {r marray code, echo=FALSE}
# Function to create a m-array based on capture-histories (CH)
marray <- function(CH){
  nind <- dim(CH)[1]
  n.occasions <- dim(CH)[2]
  m.array <- matrix(data = 0, ncol = n.occasions+1, nrow = n.occasions)
  # Calculate the number of released individuals at each time period
  for (t in 1:n.occasions){
    m.array[t,1] <- sum(CH[,t])
  }
  for (i in 1:nind){
    pos <- which(CH[i,]!=0)
    g <- length(pos)
    for (z in 1:(g-1)){
      m.array[pos[z],pos[z+1]] <- m.array[pos[z],pos[z+1]] + 1
    } #z
  } #i
  # Calculate the number of individuals that is never recaptured
  for (t in 1:n.occasions){
    m.array[t,n.occasions+1] <- m.array[t,1] - sum(m.array[t,2:n.occasions])
  }
  out <- m.array[1:(n.occasions-1),2:(n.occasions+1)]
  return(out)
}
```

``` {r ALN M-Array}

ALN_adults <- adults[(adults$site == "ALN"),]

ALN_adults_CH <- data.frame(individuals = c(unique(ALN_adults$ring)))

ALN_years <- sort(unique(ALN_adults$year))

for (year in ALN_years) {
  ALN_adults_CH[[as.character(year)]] <- NA
}

for (i in 1:nrow(ALN_adults_CH)) {
  individual <- ALN_adults_CH$individuals[i]
  capture_years <- unique(ALN_adults$year[ALN_adults$ring == individual])
  ALN_adults_CH[i, as.character(capture_years)] <- 1
}

ALN_adults_CH[is.na(ALN_adults_CH)] <- 0
ALN_adults_CH_matrix <- as.matrix(ALN_adults_CH[,-1])
ALN_adults_CH_matrix <- (apply(ALN_adults_CH_matrix, 2, as.numeric))

ALN_adults_CH_matrix

ALN_adults_marray <- marray(ALN_adults_CH_matrix)

ALN_adults_marray
```

The M-array above, generated from capture histories above the m-array using Kery and Schaub's marray function, shows individual capture histories at site 'ALN' collapsed into, on the diagonal, individuals marked and released on the nth occasion and recaptured on the n+1th occasion. The next column over on the same row shows individuals released on the nth occasion and recaptured on the n+2nd occasion. The far right column represents individuals marked and never recaptured. This is the format the IPM will take for the CMR portion of the model to estimate apparent survival. 